---
title: "STAA 567 Assignment 5"
author: "Jon Dollard"
date: "11/25/2020"
output: pdf_document
---

```{r setup, include=FALSE, fig.show='hold', out.width="50%"}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(gridExtra) #gives you grid.arrange()
library(knitr)     #making tables
library(ggallin)   #pseudolog x axis transformation
```

1.  The Cauchy distribution is used in extreme value theory.  Extreme value theory is used to model events that occur infrequently, but which involve large deviations from the 'norm'.  The standard Cauchy density is given by

$f(x)\ = \frac{1}{\pi (1 + x^2)}$, $-\infty < x < \infty$.

The cumulative distribution function for the standard Cauchy is

$F(x)\ = \frac{1}{\pi}arctan(x) + \frac{1}{2}$.

Note:  The mean and variance of the Cauchy distribution are both undefined.

Included your R code for each of the following:

(a)  Use the inverse CDF method to generate a random sample of 10,000 observations from the standard Cauchy distribution.  Plot a histogram of your sample using ggplot.  What striking features do you notice about your histogram?

```{r, warning=FALSE}
n <- 10000
u <- runif(n)
x <- tan(pi * (u - 0.5))

x_df = as.data.frame(x)
ggplot(x_df, aes(x = x)) +
 geom_histogram(aes(y = ..density..), color = "red", fill = "green", bins = 75, alpha = 0.5) +
  scale_x_continuous(trans=pseudolog10_trans)
  
```

The most striking thing that I notice about this histogram are the extremly long tails to account for the infrequently occuring large and small values generated by the distribution.  In fact, we need to rescale the x axis to a log scale in order to see the shape of the histogram (without rescaling it just looks like a single rectangle).

(b)  Repeat (a) three times (so you'll have 4 histograms between parts a and b).  Are all of your histograms similar?  Would you expect it to?  Why, or why not?  Note:  Use the same axes for your graphs from part a and b.

```{r, warning=FALSE}
#1st Repetition
n <- 10000
u <- runif(n)
x <- tan(pi * (u - 0.5))

x_df = as.data.frame(x)
plot1 <- ggplot(x_df, aes(x = x)) +
 geom_histogram(aes(y = ..density..), color = "red", fill = "green", bins = 75) +
  scale_x_continuous(trans=pseudolog10_trans)

#2nd Repetition
n <- 10000
u <- runif(n)
x <- tan(pi * (u - 0.5))

x_df = as.data.frame(x)
plot2 <- ggplot(x_df, aes(x = x)) +
 geom_histogram(aes(y = ..density..), color = "red", fill = "green", bins = 75) +
  scale_x_continuous(trans=pseudolog10_trans)

#3rd Repetition
n <- 10000
u <- runif(n)
x <- tan(pi * (u - 0.5))

x_df = as.data.frame(x)
plot3 <- ggplot(x_df, aes(x = x)) +
 geom_histogram(aes(y = ..density..), color = "red", fill = "green", bins = 75) +
  scale_x_continuous(trans=pseudolog10_trans)

grid.arrange(plot1, plot2, plot3, ncol = 1)
```

All of the histograms do have a generally similar shape with the main difference being how long the tails are (how far they extend from either side of the 'norm').  This is to be expected since the Cauchy distribution generates most of the values near the 'norm' with a few infrequently occuring values with large deviations from the 'norm'.

\pagebreak

2.  In this problem we consider simulating from the logistic distribution using the rejection method.  A logistic distribution is similar to the normal distribution in shape, but with heavier tails.  This distribution is used in logistic regression, but also in other applications, such as modeling river discharge or rainfall totals.  The logistic density is given by

$f(x)\ = \frac{exp(-x)}{(1\ +\ exp(-x))^2}$, $-\infty < x < \infty$.

Consider for now only the interval (0, $\infty$), because exp(-x) > 0 (and therefore 1 + exp(-x) > 1), we have f(x) $\leq$ exp(-x), which is the density of a standard exponential distribution Exp(1).  Positive values from the logistic distribution can be simulated using the accept-rejection algorithm with the exponential density as the envelope function.

To allow for negative values, we add an extra step to the previous rejection sampling process: if you accept a (positive) value for x for the sample, then generate an additional value

s\ = -1, with probability 0.5 and 1, with probability 0.5.

and define $y\ = s \times x$ as your simulated value.

(a) Plot the logistic distribution with the envelope superimposed on the plot (the Exp(1) distribution).  Based on the plot, is the Exp(1) a good envelope when used with the $y\ = s \times x$ method described above?

```{r, warning=FALSE}
#Set a sequence of values for x
x_grid <- seq(from = 0, to = 10, length.out = 1000)

#Create a vector for the logistic distribution
f_x = function(x){
  exp(-x) / ((1 + exp(-x))^2)
}
  
#Create a vector for the envelope distribution
e_x = function(x){
  dexp(x, rate = 1)
}

#Create a data frame to use with ggplot
plot_df <- as.data.frame(x_grid, f_x(x_grid), e_x(x_grid))

#Create a plot of the logistic distribution and envelope
ggplot(data = plot_df, aes(x = x_grid, y = e_x(x_grid))) +
  geom_line(color = 'red') +
    geom_line(aes(x = x_grid, y = f_x(x_grid)), color = 'green') +
      labs(x = "x", y = "Exp(1) and f(x)", 
         title = "Exponential and Logistic Densities vs. x") 
```

Exp(1) is a good envelope when used with the $y\ = s \times x$ method.  It does have a large reject region as x approaches zero so there may be better options to improve algorithm efficiency by investigating other densities to use as the envelope.

(b) Use the method described above to generate a sample of size 1000 from the logistic distribution.  Provide a histogram for your simulated data.  Report the acceptance rate.

```{r, warning=FALSE}
set.seed(1)
n_samps <- 1000
n_accepts <- 0
n_rejects <- 0
samp_exp <- numeric(n_samps)
#s <- 0
#y <- 0

while(n_accepts < n_samps){
  
      y <- rexp(1)
      u <- runif(1)
    
      #Check if we are in the accept region
      if (u < f_x(y)/e_x(y)){
        #In the accept region so increment the accept counter
        n_accepts <- n_accepts + 1
        #take a sample (-1,1) of size 1 with a 50/50 probability
        s <- sample(c(-1,1), size = 1, replace = TRUE, prob = c(0.5,0.5))
        #keep the accept value y in the samp vector at the n_accept location in the vector
        #after adjusting the sign
        samp_exp[n_accepts] = y * s
      }
    
      #Not in the accept region so increment the reject counter and go back
      #to the if statement
      else{
        n_rejects <- n_rejects + 1
      }
  
}

#Plot a histogram of the simulated logistic distribution
logist_data <- as.data.frame(samp_exp)
ggplot(data = logist_data, aes(x = samp_exp)) +
 geom_histogram(aes(y = ..density..), color = "green", fill = "red", alpha = 0.4, bins = 50) + 
  xlim(c(-8,8))

#Calculate acceptance rate
accept_rate <- n_accepts / (n_accepts + n_rejects)
accept_rate

min(samp_exp)

```

```{r}
set.seed(1)
n.samp = 1000
n = 0
n_all = 0
samps = numeric(n.samp)
while(n < n.samp){
  n_all = n_all + 1
  y = rexp(1)
  u = runif(1)
  if(u < f_x(y)/e_x(y)){
    n = n+1
    s = sample(c(-1,1), 1)
    samps[n] = s * y
  }
}
n/n_all
min(samps)
```


(c)  Now consider using the standard normal distribution N(0,1) as the envelope.  Repeat part (a) using N(0,1) as the envelope.

```{r, warning=FALSE}
#Set a sequence of values for x
x_grid <- seq(from = -10, to = 10, length.out = 1000)

#Create a vector for the envelope distribution
e_x = function(x){
   dnorm(x)
}

#Create a data frame to use with ggplot
plot_df <- as.data.frame(x_grid, f_x(x_grid), e_x(x_grid))

#Create a plot of the logistic distribution and envelope
ggplot(data = plot_df, aes(x = x_grid, y = e_x(x_grid))) +
  geom_line(color = 'red') +
    geom_line(aes(x = x_grid, y = f_x(x_grid)), color = 'green') +
      labs(x = "x", y = "e(x) and f(x)", 
         title = "Normal and Logistic Densities vs. x") 
```

The standard normal distribution does not appear to be a good envelope for the logistic distribution because it isn't larger than the logistic distribution over the support.  We can see from the plot that if we use the standard normal distribution as the envelope we won't sample the tails of the logistic distribution.

(d) Using N(0,1) as the envelope, generate a sample of size 1000 from the logistic distribution.  Provide a histogram for your simulated data.  Report the acceptance rate.

```{r, warning=FALSE}
set.seed(567)
n_samps <- 1000
n_accepts <- 0
n_rejects <- 0
samp_norm <- numeric(n_samps)
y <- 0

while(n_accepts < n_samps){
  
      y <- rnorm(1)
      u <- runif(1)
    
      #Check if we are in the accept region
      if (u < f_x(y)/e_x(y)){
        #In the accept region so increment the accept counter
        n_accepts <- n_accepts + 1
        samp_norm[n_accepts] = y 
      }
    
      #Not in the accept region so increment the reject counter and go back
      #to the if statement
      else{
        n_rejects <- n_rejects + 1
      }
  
}

#Plot a histogram of the simulated logistic distribution
logist_data <- as.data.frame(samp_norm)
ggplot(data = logist_data, aes(x = samp_norm)) +
 geom_histogram(aes(y = ..density..), color = "green", fill = "red", alpha = 0.4, bins = 50) + 
  xlim(c(-8,8))

#Calculate acceptance rate
accept_rate <- n_accepts / (n_accepts + n_rejects)
accept_rate
```

(e) Simulate 1000 draws from the logistic distribution using the rlogis command in R.  Create a table of the summary statistics of your simulation from parts b, d, and e.  (Use the summary command).

Discuss:  Is the method you used in part b or part d more appropriate?  Why?

```{r}
set.seed(1)
n <- 1000
logis_dist_samp <- rlogis(n)

summary_exp <- round(summary(samp_exp), 3)
summary_exp
summary_norm <- round(summary(samp_norm), 3)
summary_norm
summary_logis <- round(summary(logis_dist_samp), 3)
summary_logis
envelope <- matrix(c("Exp(1)", "N(0,1)", "rlogis()"))

summary_matrix <- cbind(envelope, matrix(c(summary_exp, summary_norm, summary_logis),
                                         nrow = 3, byrow = TRUE))

kable(summary_matrix, col.names = c("Envelope", "Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max."),
      align = "ccc", caption = 
        "Simulation Summary Statistics")

```

The method we used in part (b) is more appropriate.  The Exp(1) envelope doesn't have a high acceptance rate and might not be the most efficient envelope, but it does meet all the requirements for being an envelope.  The standard normal distribution is not an adequate envelope because it is not larger than the sample distribution over the support.  We can see from the histogram of sampled values in part (d) that we don't capture the tails of the logistic distribution when using the standard normal envelope.






























































