---
title: "STAA 567 - Assignment 1"
author: "Jon Dollard"
date: "10/23/2020"
output: pdf_document
---

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(gridExtra) #gives you grid.arrange()
library(pracma)    #gives you the movavg() function
library(knitr)     #making tables
library(lubridate) 
```

COVID-19 data: In this problem you’ll examine the number of positive COVID-19 tests in Larimer
County, Colorado. For each day, you’ll need to compute the 14-day running average of COVID-19 cases.
You’ll consider multiple approaches to do the computation.

Use the COVID-19 data to do the following:

 - Method 1: Use a for loop. Create a function for your for loop with arguments that allow
   you to change the raw data (here cases per day) and length of running average (here 14).
   
 - Method 2: Create a function that uses the cumsum function instead of a for loop to compute the
   running average.
   
 - Method 3: Use another method. Design your own or find one in a package. 
 

```{r, message=FALSE}
#Read in the data to r
cases = read_csv("2020.10.22.Larimer_cases-by-day.csv")

#Look at a summary of cases to make sure things look right
head(cases)
summary(cases)
length(cases$Cases)
```

```{r}
#Use a for loop to calculate the 14-day running average
#create a variable for the 14-day running average and initialize it outside the for loop to NULL
average_14_day_fl = NULL

for(i in 1:(length(cases$Cases)-13)){
  average_14_day_fl[i] = sum(cases$Cases[i:(i+13)])/14
}

#take a look at the first 5 values of the average 14 day running average
head(average_14_day_fl)
```

```{r}
#For method 2 we will use the cumsum function to calculate the 14 day running average
n = 14

#create a vector with the cases values in them
cases_vec <- cases$Cases

#Create a vector with cumulative sums 
cum_cases <- c(0,cumsum(cases_vec))

#Now we can calculate the 14 day running averages
average_14_day_cs = (cum_cases[(n+1):length(cum_cases)] - cum_cases[1:(length(cum_cases)-n)])/n

#Take a look at the first 5 values
head(average_14_day_cs)

#Quick check for equivalence with the 14-day running average using the for loop
identical(average_14_day_fl,average_14_day_cs)
```

```{r}
#For method 3 I will use movavg function in the pracma library
average_14_day_mo = movavg(cases$Cases, 14, type = "s")

#Remove the first 13 days from the vector
average_14_day_mo <- average_14_day_mo[14:length(average_14_day_mo)]

#Take a look at the first 5 values
head(average_14_day_mo)

#Quick check for equivalence with the other methods
identical(average_14_day_cs,average_14_day_mo)
```

To turn in:

1. Make a table to show the running average for the last 5 days computed using all 3 methods. If
you’ve done things correctly, all three columns will be identical.

```{r}
#create a table of the last 5 days computed using all 3 methods
table_matrix <- matrix(c((average_14_day_fl[(length(average_14_day_fl)-4):length(average_14_day_fl)])                                                  ,(average_14_day_cs[(length(average_14_day_cs)-4):length(average_14_day_cs)]),                                                  (average_14_day_mo[(length(average_14_day_mo)-4):length(average_14_day_mo)])),
                         nrow = 5, ncol = 3, 
                         dimnames = list(c(cases$Date[(length(cases$Date)-4):length(cases$Date)])))

kable(table_matrix, col.names = c("Method 1", "Method 2", "Method 3"), align = "ccc", caption = "14 Day Running Average Using 3 Methods")
```


2. Plot your output to show the 14-day running average over time computed using all the methods
that you created. Use the facet_grid command in ggplot or some other approach to show the
results for all 3 methods.

```{r}
#Create data frames for each method to use with ggplot
meth1_df <- data.frame(y = average_14_day_fl, x = cases$Date[14:length(cases$Date)])
meth2_df <- data.frame(y = average_14_day_cs, x = cases$Date[14:length(cases$Date)])
meth3_df <- data.frame(y = average_14_day_mo, x = cases$Date[14:length(cases$Date)])

#Use ggplot to plot the 3 methods on different plots
meth1_plot <- ggplot() +
              geom_line(data = meth1_df, aes(x = mdy(x), y = y)) 
meth2_plot <- ggplot() +
              geom_line(data = meth2_df, aes(x = mdy(x), y = y))
meth3_plot <- ggplot() +
              geom_line(data = meth3_df, aes(x = mdy(x), y = y))

grid.arrange(meth1_plot, meth2_plot, meth3_plot)
```

3. Compare the timing of your methods. To get a meaningful comparison, replicate the data 10,000
times so that you have a larger data set. Report the times for your methods.

```{r}
#calculate system times for each method and the discuss

#For method 1 create a function using with a for loop
method1 = function(x, n){
  average_14_day_fl = NULL  
      for(i in 1:(length(x)-13)){
      average_14_day_fl[i] = sum(x[i:(i+13)])/14
      }
    }

#Calculate system time for 10,000 repetitions
sys_time_m1 <- system.time(method1(rep(cases$Cases,10000),14))

#Report system time for method 1
sys_time_m1

#For method 2 create a function using the cumsum function
method2 = function(x,n){
  cases_vec <- x
  cum_cases <- c(0,cumsum(x))
  average_14_day_cs = (cum_cases[(n+1):length(cum_cases)] - cum_cases[1:(length(cum_cases)-n)])/n
}

#Calculate system time for 10,000 repetitions
sys_time_m2 <- system.time(method2(rep(cases$Cases,10000),14)) 

#Report system time for method 2
sys_time_m2

#For method 3 create a function using the movavg() function in the pracma library
method3 = function(x,n){
  average_14_day_mo = movavg(x, n, type = "s")
}

#Calculate the system time for 10,000 repetitions
sys_time_m3 <- system.time(method3(rep(cases$Cases,10000),14)) 

#Report the system time for method 3
sys_time_m3
```






































