---
title: "STAA 567 - Assignment 1"
author: "Jon Dollard"
date: "10/23/2020"
output: pdf_document
---

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(gridExtra) #gives you grid.arrange()
library(pracma)    #gives you the movavg() function
library(knitr)     #making tables
library(lubridate) 
```

COVID-19 data: In this problem you’ll examine the number of positive COVID-19 tests in Larimer
County, Colorado. For each day, you’ll need to compute the 14-day running average of COVID-19 cases.
You’ll consider multiple approaches to do the computation.

Use the COVID-19 data to do the following:

 - Method 1: Use a for loop. Create a function for your for loop with arguments that allow
   you to change the raw data (here cases per day) and length of running average (here 14).
   
 - Method 2: Create a function that uses the cumsum function instead of a for loop to compute the
   running average.
   
 - Method 3: Use another method. Design your own or find one in a package. 
 

```{r, message=FALSE}
#Read in the data to r
cases = read_csv("2020.10.22.Larimer_cases-by-day.csv")

#Look at a summary of cases to make sure things look right
head(cases)
summary(cases)
length(cases$Cases)
```

```{r}
#Use a for loop to calculate the 14-day running average
x = cases$Cases
n = 14
average_14_day_fl = NULL

#For method 1 create a function using with a for loop
method1 = function(x, n){
      for(i in 1:(length(x)-(n-1))){
      average_14_day_fl[i] = sum(x[i:(i+(n-1))])/n
      }
      return(average_14_day_fl)
    }

meth1_RA <- method1(x,n)

#take a look at the first 5 values of the average 14 day running average
head(meth1_RA)
```

```{r}
#For method 2 create a function using the cumsum function
method2 = function(x,n){
  cases_vec <- x
  cum_cases <- c(0,cumsum(x))
  average_14_day_cs = (cum_cases[(n+1):length(cum_cases)] - cum_cases[1:(length(cum_cases)-n)])/n
  return(average_14_day_cs)
}

#Take a look at the first 5 values
meth2_RA <- method2(x,n)
head(meth2_RA)

#Quick check for equivalence with the 14-day running average using the for loop
identical(meth1_RA, meth2_RA)
```

```{r}
#For method 3 create a function using the movavg() function in the pracma library
method3 = function(x,n){
  average_14_day_mo = movavg(x, n, type = "s")
  return(average_14_day_mo[n:length(x)])
}

#Take a look at the first 5 values
meth3_RA <- method3(x,n)
head(meth3_RA)

#Quick check for equivalence with the other methods
identical(meth2_RA, meth3_RA)
```

To turn in:

1. Make a table to show the running average for the last 5 days computed using all 3 methods. If
you’ve done things correctly, all three columns will be identical.

```{r}
#create a table of the last 5 days computed using all 3 methods
table_matrix <- matrix(c((meth1_RA[(length(meth1_RA)-4):length(meth1_RA)]),                                                                             (meth2_RA[(length(meth2_RA)-4):length(meth2_RA)]),                                                                             (meth3_RA[(length(meth3_RA)-4):length(meth3_RA)])),
                         nrow = 5, ncol = 3, 
                         dimnames = list(c(cases$Date[(length(cases$Date)-4):length(cases$Date)])))

kable(table_matrix, col.names = c("Method 1", "Method 2", "Method 3"), align = "ccc", caption = "14 Day Running Average Using 3 Methods")
```


2. Plot your output to show the 14-day running average over time computed using all the methods
that you created. Use the facet_grid command in ggplot or some other approach to show the
results for all 3 methods.

```{r}
#Create data frames for each method to use with ggplot
meth1_df <- data.frame(y = meth1_RA, x = cases$Date[n:length(cases$Date)])
meth2_df <- data.frame(y = meth2_RA, x = cases$Date[n:length(cases$Date)])
meth3_df <- data.frame(y = meth3_RA, x = cases$Date[n:length(cases$Date)])

#Use ggplot to plot the 3 methods on different plots
meth1_plot <- ggplot() +
              geom_line(data = meth1_df, aes(x = mdy(x), y = y), color = "red") +
              labs(x = "Months", y = "14-day Running Avg", title = "14-day Running Average Using Method 1 - For Looping") +
              theme(plot.title = element_text(hjust = 0.5))
meth2_plot <- ggplot() +
              geom_line(data = meth2_df, aes(x = mdy(x), y = y), color = "blue") +
              labs(x = "Months", y = "14-day Running Avg", 
              title = "14-day Running Average Using Method 2 - cumsum() Function") +               
              theme(plot.title = element_text(hjust = 0.5))
meth3_plot <- ggplot() +
              geom_line(data = meth3_df, aes(x = mdy(x), y = y), color = "green") +
              labs(x = "Months", y = "14-day Running Avg", 
              title = "14-day Running Average Using Method 3 - movavg() Function") +
              theme(plot.title = element_text(hjust = 0.5))

grid.arrange(meth1_plot, meth2_plot, meth3_plot)
```

3. Compare the timing of your methods. To get a meaningful comparison, replicate the data 10,000
times so that you have a larger data set. Report the times for your methods.

```{r}
#calculate system times for each method and the discuss

#Calculate system time for 10,000 repetitions
sys_time_m1 <- system.time(method1(rep(cases$Cases,10000),14))

#Report system time for method 1
sys_time_m1

#Calculate system time for 10,000 repetitions
sys_time_m2 <- system.time(method2(rep(cases$Cases,10000),14)) 

#Report system time for method 2
sys_time_m2

#Calculate the system time for 10,000 repetitions
sys_time_m3 <- system.time(method3(rep(cases$Cases,10000),14)) 

#Report the system time for method 3
sys_time_m3
```

It looks like for the 3 methods I used that the cumsum() is the fastest performing.  The for loop function is next followed by the movavg() function in the pracma library which seems very slow in comparison.  After reviewing the github code for the movavg() we can see why this function is so slow as it allows for multiple ways of calculating the running average.  The function is relatively long in comparison to the other two methods with numerous conditional branches depending upon parameter settings of the function.  Very interesting though to dig further into 




































